---
title: "Analyzing CEO Compensation: Correlations & Predictive Modeling"
author: "Anushna Gunda, Inaya Rizvi, Harsh Malik, Sai Bathina"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
    number_sections: yes
    theme: united
    code_folding: none
---


Below I have created first-level headings for the project. You can add your own second-level (or third-level) headings by adding ## or ### before the sub-heading. I also added a code chunk under the #Results section just to show you how to get started. Once you are ready to knit your markdown file into a Word doc, delete this paragraph. kl

# Introduction
CEO compensation is a key metric in corporate governance and financial analysis. Our goal is to analyze what factors are most correlated with CEO pay and whether we can predict CEO compensation based on financial and market indicators.

# Research Questions
- What company financial and market factors influence CEO compensation?
- Can we create a regression model to predict CEO total compensation?

# Data and Methods

**Source:** Pitchbook 

## Overview of Dataset
Using pitchbook, we custom-formatted a dataset that includes **CEO total compensation** as the dependent variable and the following independent variables:

- **Company Financials:** Total Revenue, Revenue per Employee, EBITDA Margin, etc.

- **Profitability Metrics:** ROIC, Net Profit Margin, Revenue Growth.

- **Debt & Leverage:** Net Debt, Debt to Capital, Total Assets.

- **Market Indicators:** Price % Change YTD, Beta (5-year), EV, EV/EBITDA.

- **Categorical Factors:** HQ State/Province, Primary Industry Group (for identification and clustering)

## Data Import & Preprocessing
Includes formatting column names and negative values (ceo_data), and handling null values by replacing null with medians (quantitative) or mode (qualitative) (imputed_data)

AFTER RUNNING THIS CODE CHUNK:
there are 2 datasets you can use:

ceo_data = clean but with NA values for missing data
imputed_data = clean with NA values replaced with median or mode (depending on data type) 

```{r, message=FALSE, warning=FALSE}

# Loading Required Libraries
library(tidyverse)   # Data manipulation
library(mice)        # Missing data imputation
library(dplyr)

# Loading the Dataset
raw_data <- read.csv("ceo_comp_data.csv", na.strings = c("", "NA"))
raw_data <- raw_data %>% select (-X, -X.1)

# Fixing Column Names
ceo_data <- raw_data %>%
  rename(
    total_revenue = "Total.Revenue..FY.",
    revenue_per_employee= "Revenue.per.Employee..FY.", 
    roic = "ROIC..Return.on.Invested.Capital...FY.",
    net_profit_margin = "Net.Profit.Margin..FY.",
    ebitda_margin = "EBITDA.Margin..FY.",
    revenue_growth = "Revenue...Growth..FY.",
    net_debt = "Net.Debt..FY.",
    debt_to_capital = "Debt.to.Capital..FY.",
    total_assets = "Total.Assets..FY.",
    price_change_ytd = "Price...Change.YTD",
    ev = "EV..FY.",
    hq = "HQ.State.Province",
    primary_industry_group = "Primary.Industry.Group",
    ceo = "CEO",
    ceo_total_comp_mils = "CEO.Total.Compensation..in.millions.",
    employees = "Employees",
    companies = "Companies"
  )

# Viewing the new column names
colnames(ceo_data)

# Viewing Basic Info
str(ceo_data)   # Check structure
summary(ceo_data)  # Summary statistics
print(ceo_data)  # Quick data preview

# Function to clean and convert numeric columns with commas & parentheses
convert_numeric_column <- function(column) {
  column <- as.character(column)  # Ensure it's a character string first
  
  # Remove any leading/trailing spaces
  column <- trimws(column)
  
  # Convert negative values (e.g., "(2,000)" → "-2000")
  column <- gsub("[(),]", "", column)  # Remove parentheses and commas
  column <- ifelse(grepl("^\\(.*\\)$", column), paste0("-", column), column) 
  
  # Convert to numeric
  column <- as.numeric(column)
  
  return(column)
}

# Apply conversion to `total_revenue` & `total_assets`
ceo_data$total_revenue <- convert_numeric_column(ceo_data$total_revenue)
ceo_data$total_assets <- convert_numeric_column(ceo_data$total_assets)
ceo_data$employees <- convert_numeric_column(ceo_data$employees)

# Check if conversion worked
str(ceo_data$total_revenue)
str(ceo_data$total_assets)


numeric_cols <- c("roic", "net_profit_margin", "ebitda_margin", "revenue_growth", 
                   "net_debt", "price_change_ytd", "ev", "revenue_per_employee",
                  "debt_to_capital", "ceo_total_comp_mils")

# Defining function to properly format negative numeric values 
convert_negatives_column <- function(column) {
  column <- as.character(column)  # Convert to character
  
  for (i in seq_along(column)) {  
    if (!is.na(column[i]) && column[i] != "") {  # Skip NA or empty values
      column[i] <- trimws(column[i])  # Remove any leading/trailing spaces
      
      if (grepl("^\\(.*\\)$", column[i])) {  # Check if value has parentheses
        column[i] <- gsub("[(),]", "", column[i])  # Remove parentheses & commas
        column[i] <- as.numeric(column[i]) * -1  # Convert to negative
      } else {
        column[i] <- gsub(",", "", column[i])  # Remove commas from numbers
        column[i] <- as.numeric(column[i])  # Convert normally
      }
      
      # If conversion fails, set it to NA explicitly (to catch errors)
      if (is.na(column[i])) {
        column[i] <- NA
      }
    }
  }
  
  return(as.numeric(column))  # Ensure the final output is numeric
}


# Apply fix to all numeric columns
for (col in numeric_cols) {
  ceo_data[[col]] <- convert_negatives_column(ceo_data[[col]])
}


#  Create a Copy for Imputation
imputed_data <- ceo_data

#  Defining function to impute numeric columns with median
impute_median <- function(column) {
  column[is.na(column)] <- median(column, na.rm = TRUE)
  return(column)
}

# Imputing numeric columns
imputed_data[numeric_cols] <- lapply(imputed_data[numeric_cols], impute_median)


# Defining function to impute categorical columns with mode
mode_impute <- function(column) {
  column[is.na(column)] <- names(sort(table(column), decreasing = TRUE))[1]
  return(column)
}

# Imputing categorical columns 
categorical_cols <- c("primary_industry_group", "hq")  
imputed_data[categorical_cols] <- lapply(imputed_data[categorical_cols], mode_impute)


# Ensuring that all numeric columns are actually numeric type for both 
ceo_data[numeric_cols] <- lapply(ceo_data[numeric_cols], function(x) as.numeric(as.character(x)))
imputed_data[numeric_cols] <- lapply(imputed_data[numeric_cols], function(x) as.numeric(as.character(x)))


# Check the Cleaned Data
summary(ceo_data)
glimpse(ceo_data)

# Check the Imputed Dataset
summary(imputed_data)
glimpse(imputed_data)

#  Optional - Save Cleaned Datasets
# write.csv(ceo_data, "ceo_comp_data_raw.csv", row.names = FALSE)
# write.csv(imputed_data, "ceo_comp_data_imputed.csv", row.names = FALSE)

```


## Exploratory Analysis

### Summary Statistics
```{r}

```

### Distributions of key variables 
-- maybe histogram of CEO compensation distribution
```{r}

```

### Correlation matrix to check relationships between dependent variables
```{r}

```
-- interpret correlations, are higher revenues associated with higher CEO pay? or debt levels, etc. 

### Visualizations 
some ideas maybe:
-- box plots of CEO comp by industry (also maybe run ANOVA test in conjunction to see if CEO comp differs significantly between industries)
-- scatter plots of CEO comp vs. total rev, ROIC, net income growth
-- test CEO compensation differences based on debt-levels 
      -- split companies into high and low debt using median as             divider
      -- use a t-test (if normal dist.), Wilcoxon test (if skewed)             (look at overall histogram of CEO comp for this)
```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(car)
library(stats)

# Using the imputed_data dataset from the provided information

# 1. Box plots of CEO compensation by industry with ANOVA test
# First, let's get the top industries by frequency to avoid too many categories
top_industries <- imputed_data %>%
  count(primary_industry_group, sort = TRUE) %>%
  filter(n >= 10) %>% # filter for industries with at least 10 companies
  pull(primary_industry_group)

# Create a filtered dataset with only top industries
industry_filtered <- imputed_data %>%
  filter(primary_industry_group %in% top_industries)

# Create boxplot of CEO compensation by industry
ggplot(industry_filtered, aes(x = reorder(primary_industry_group, ceo_total_comp_mils, FUN = median), 
                              y = ceo_total_comp_mils)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "CEO Compensation by Industry",
       x = "Industry Group",
       y = "CEO Total Compensation (Millions $)")

# Run ANOVA test to check if CEO compensation differs significantly between industries
anova_result <- aov(ceo_total_comp_mils ~ primary_industry_group, data = industry_filtered)
summary(anova_result)

# 2. Scatter plots of CEO comp vs. financial metrics

# CEO comp vs Total Revenue
ggplot(imputed_data, aes(x = total_revenue, y = ceo_total_comp_mils)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  scale_x_log10(labels = scales::comma) +
  theme_minimal() +
  labs(title = "CEO Compensation vs. Total Revenue",
       x = "Total Revenue (log scale)",
       y = "CEO Total Compensation (Millions $)")

# Calculate correlation
cor_rev_comp <- cor.test(log10(imputed_data$total_revenue + 1), imputed_data$ceo_total_comp_mils)
print(cor_rev_comp)

# CEO comp vs ROIC
ggplot(imputed_data, aes(x = roic, y = ceo_total_comp_mils)) +
  geom_point(alpha = 0.6, color = "green") +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "CEO Compensation vs. Return on Invested Capital",
       x = "ROIC (%)",
       y = "CEO Total Compensation (Millions $)") +
  coord_cartesian(xlim = c(-100, 100)) # Limit x-axis to exclude extreme outliers

# Calculate correlation (excluding extreme outliers)
roic_filtered <- imputed_data %>% filter(roic > -100 & roic < 100)
cor_roic_comp <- cor.test(roic_filtered$roic, roic_filtered$ceo_total_comp_mils)
print(cor_roic_comp)

# CEO comp vs Revenue Growth
ggplot(imputed_data, aes(x = revenue_growth, y = ceo_total_comp_mils)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(title = "CEO Compensation vs. Revenue Growth",
       x = "Revenue Growth (%)",
       y = "CEO Total Compensation (Millions $)") +
  coord_cartesian(xlim = c(-50, 100)) # Limit x-axis to exclude extreme outliers

# Calculate correlation (excluding extreme outliers)
growth_filtered <- imputed_data %>% filter(revenue_growth > -50 & revenue_growth < 100)
cor_growth_comp <- cor.test(growth_filtered$revenue_growth, growth_filtered$ceo_total_comp_mils)
print(cor_growth_comp)

# 3. Test CEO compensation differences based on debt-levels
# First, check the histogram of CEO compensation to determine if normal or skewed
ggplot(imputed_data, aes(x = ceo_total_comp_mils)) +
  geom_histogram(bins = 30, fill = "orange", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of CEO Compensation",
       x = "CEO Total Compensation (Millions $)",
       y = "Frequency")

# Split companies into high and low debt using median as divider
debt_median <- median(imputed_data$debt_to_capital, na.rm = TRUE)
imputed_data$debt_level <- ifelse(imputed_data$debt_to_capital > debt_median, "High Debt", "Low Debt")

# Visualize CEO compensation by debt level
ggplot(imputed_data, aes(x = debt_level, y = ceo_total_comp_mils)) +
  geom_boxplot(fill = c("lightgreen", "coral"), alpha = 0.7) +
  theme_minimal() +
  labs(title = "CEO Compensation by Debt Level",
       x = "Debt Level",
       y = "CEO Total Compensation (Millions $)")

# Test for normality
shapiro.test(imputed_data$ceo_total_comp_mils[1:5000]) # Limited to 5000 samples due to Shapiro-Wilk test limitations

# Based on histogram and normality test, choose appropriate test
# If skewed (likely), use Wilcoxon test
wilcox_result <- wilcox.test(ceo_total_comp_mils ~ debt_level, data = imputed_data)
print(wilcox_result)

# If normal distribution (unlikely), use t-test
t_test_result <- t.test(ceo_total_comp_mils ~ debt_level, data = imputed_data)
print(t_test_result)

# 4. Additional analysis: Company size (employees) vs CEO compensation
ggplot(imputed_data, aes(x = employees, y = ceo_total_comp_mils)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red") +
  scale_x_log10(labels = scales::comma) +
  theme_minimal() +
  labs(title = "CEO Compensation vs. Number of Employees",
       x = "Number of Employees (log scale)",
       y = "CEO Total Compensation (Millions $)")

# Calculate correlation
cor_emp_comp <- cor.test(log10(imputed_data$employees + 1), imputed_data$ceo_total_comp_mils)
print(cor_emp_comp)
## Regression Analysis
```

### Multiple Linear Regression: CEO Compensation ~ Financial Metrics

-- which variables are statistically significant?
-- are there unexpected relationships?
-- check for multicollinearity
-- use R-squared and residual plots to assess model performance



-- consider running a Shapiro-Wilk and Breusch-Pagan Test to check assumptions in model (ask inaya or chatGPT if you have questions)


### Network Analysis - Clustering

-- Are there industry-based clusters where CEO pay is higher?

```{r, warning=FALSE, message=FALSE}
# Load required libraries
library(ggplot2)
library(dplyr)
library(viridis)
library(stringr)
library(ggrepel)

# Create location summary statistics
location_summary <- imputed_data %>%
  filter(!is.na(hq)) %>%
  group_by(hq) %>%
  summarize(
    avg_ceo_comp = mean(ceo_total_comp_mils, na.rm = TRUE),
    median_ceo_comp = median(ceo_total_comp_mils, na.rm = TRUE),
    count = n(),
    avg_ev = mean(ev/1e6, na.rm = TRUE),
    avg_employees = mean(employees, na.rm = TRUE)
  ) %>%
  # Filter to include only locations with at least 5 companies
  filter(count >= 5) %>%
  arrange(desc(avg_ceo_comp))

# Print top locations by CEO compensation
print(location_summary)

# Create a bar chart showing average CEO compensation by location
p_location_bar <- ggplot(location_summary, 
                         aes(x = reorder(hq, avg_ceo_comp), 
                             y = avg_ceo_comp, 
                             fill = avg_ceo_comp)) +
  geom_col() +
  geom_text(aes(label = sprintf("$%.1fM", avg_ceo_comp)), 
            hjust = -0.1, size = 3) +
  geom_text(aes(label = sprintf("n=%d", count)),
            y = 1, hjust = 1.1, size = 3, color = "white") +
  coord_flip() +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Average CEO Compensation by Location",
    subtitle = "For locations with at least 5 companies",
    x = NULL,
    y = "Average CEO Compensation ($ Millions)",
    fill = "Avg CEO Comp ($M)"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 9)
  )

print(p_location_bar)

# Create a scatter plot comparing CEO compensation vs. enterprise value by location
p_location_scatter <- ggplot(imputed_data %>% 
                              filter(hq %in% location_summary$hq), 
                            aes(x = ev/1e6, 
                                y = ceo_total_comp_mils,
                                color = hq)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_x_log10(labels = scales::comma_format()) +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(
    title = "CEO Compensation vs. Enterprise Value by Location",
    subtitle = "Log scale for EV, top locations by company count",
    x = "Enterprise Value ($ Millions, log scale)",
    y = "CEO Compensation ($ Millions)",
    color = "Headquarters"
  ) +
  theme(legend.position = "right")

print(p_location_scatter)

# Create a location comparison visualization
p_location_comparison <- ggplot(location_summary, 
                             aes(x = avg_ev, 
                                 y = avg_ceo_comp, 
                                 size = count,
                                 color = avg_ceo_comp)) +
  geom_point(alpha = 0.8) +
  geom_text_repel(aes(label = hq), 
                 size = 3,
                 box.padding = 0.5,
                 point.padding = 0.5,
                 force = 2) +
  scale_x_log10(labels = scales::comma_format()) +
  scale_size_continuous(range = c(3, 10)) +
  scale_color_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "CEO Compensation vs. Enterprise Value by Location",
    subtitle = "Bubble size indicates number of companies in each location",
    x = "Average Enterprise Value ($ Millions, log scale)",
    y = "Average CEO Compensation ($ Millions)",
    size = "Number of Companies",
    color = "Avg CEO Comp ($M)"
  )

print(p_location_comparison)

# Location boxplot to show distribution
p_location_boxplot <- ggplot(imputed_data %>% 
                              filter(hq %in% location_summary$hq), 
                            aes(x = reorder(hq, ceo_total_comp_mils, FUN = median), 
                                y = ceo_total_comp_mils,
                                fill = median(ceo_total_comp_mils))) +
  geom_boxplot() +
  coord_flip() +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Distribution of CEO Compensation by Location",
    subtitle = "For locations with at least 5 companies",
    x = NULL,
    y = "CEO Compensation ($ Millions)",
    fill = "Median CEO Comp ($M)"
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 9)
  )

print(p_location_boxplot)
```

# Results
Describe results in this section

You can add text above and below code chunks

# Discussion

